---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Loading packages

## Loading and preprocessing the data

```{r}
activity_data<-read.csv('activity.csv')
```


## What is mean total number of steps taken per day?

```{r,results='asis',echo=T}
activity_data_per_day<-tapply(activity_data$steps,activity_data$date,sum)

mean_daily_activity <-mean(activity_data_per_day,na.rm=T)
median_daily_activity<-median(activity_data_per_day,na.rm=T)

plot(activity_data_per_day,type='h',xlab='Day',ylab='Steps')
```

The average amount of steps per day is `r format(mean_daily_activity,scientific=F)`, whereas the median amounts to `r format(median_daily_activity,scientific=F)`.



## What is the average daily activity pattern?

```{r}
activity_per_interval<-tapply(activity_data$steps,activity_data$interval,mean,na.rm=T)

max(activity_per_interval,na.rm=T)

plot(unique(activity_data$interval),activity_per_interval,type='l',ylab='steps',xlab='interval')

max_activity<-names(which(max(activity_per_interval)==activity_per_interval))
```
The maximal observed activity was in the interval which started at `r max_activity`. 



## Imputing missing values

```{r}
summary_activity<-summary(activity_data) #NAs can be found in the summary, 

#it is apparent that NAs are only present in the 'steps' column of the data

NA_count<-sum(is.na(activity_data))
```
The total number of NA values is `r NA_count``

```{r}
#swapping the NA values for the mean of that given interval

mean_activity_vector<-rep(activity_per_interval ,61) 
#repilcates the mean interval steps for every day/observation
#the mean interval value will be picked form this vector, 
#to replace the NA (both are the same length, for-loop convenience)

no_NA_activity_data<-activity_data #create new cleaned data 

for (i in 1:dim(no_NA_activity_data)[1]){
  if(is.na(no_NA_activity_data$steps[i])){
    no_NA_activity_data$steps[i]<-mean_activity_vector[i]
  }
}
```

```{r,message=F}
no_NA_activity_data_per_day<-tapply(no_NA_activity_data$steps,no_NA_activity_data$date,sum)
no_NA_mean_daily_activity <-mean(no_NA_activity_data_per_day,na.rm=T)

rbind(plot(activity_data_per_day,type='h',xlab='Day',ylab='Steps',main='with missing values'),
plot(no_NA_activity_data_per_day,type='h',xlab='Day',ylab='Steps',main='without missing values'))
```

Given the NA values were only present for a given entire day, the days with data remain unaffected, whereas the days where no data was available, are now represented by a histogram bar that reflects the 60 day average. 

## Are there differences in activity patterns between weekdays and weekends?

```{r}
no_NA_activity_data$Weekdays<-weekdays(as.Date(activity_data$date))

weekdays<-unique(no_NA_activity_data$Weekdays)[1:5]
weekenddays<-unique(no_NA_activity_data$Weekdays)[6:7] 
#can be excluded, non weekdays are weekenddays by default

no_NA_activity_data$weekfactor<-as.factor(no_NA_activity_data$Weekdays%in% weekdays)
levels(no_NA_activity_data$weekfactor)<-c('weekend','weekday') #beware of level order!
```

```{r}
#splitting the data_set in two, for weekdays and weekends

activity_data_split<-split(no_NA_activity_data,no_NA_activity_data$weekfactor)

weekday_activity<-tapply(activity_data_split$weekday$steps,activity_data_split$weekday$interval,mean)

weekend_activity<-tapply(activity_data_split$weekend$steps,activity_data_split$weekend$interval,mean)

rbind(plot(unique(names(weekday_activity)),weekday_activity,type='l',xlab='interval',ylab='steps',main='weekday activity'),
      plot(unique(names(weekend_activity)),weekend_activity,type='l',xlab='interval',ylab='steps',main='weekend activity'))
```

